Parametricity for Primitive Nested Types and GADTs

### 2020-12-14 free-theorems-nesteds/20-mscs.tex
started on second half of journal version

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\Bcons}{\mathit{Bcons}}
\newcommand{\Bnil}{\mathit{Bnil}}

@@

\begin{defi}\label{def:wftypes}
The formation rules for the set $\F$ of\, {\em well-formed types
  {\color{red} over $\Gamma$ and $\Phi$?}} are

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;{\color{red}\ol{\gamma^0}},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash H}$}
\TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F {\color{red} [\ol{\gamma := H}]})\,\ol{G}$}  
\DisplayProof
\end{array}\]

\vspace*{0.04in}

\[\begin{array}{cc}
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
\end{defi}

{\color{red} Note that we use first-order substitution in the
  $\mu$-clause even though it isn't defined until the next
  page. Mutually inductive definition? The purpose of the substitution
  in the $\mu$-type clause of Def 2.1 is to ensure that variables in
  $\Phi$ are not applied to types involving variables from $\ol\alpha$
  in $F$; indeed, the first-order substitution $[\ol{\gamma := H}]$
  constrains the types in $\ol H$ to have arity $0$, and constrains
  the variables in $\Phi$ to already be fully applied to types not
  involving $\ol\alpha$ in $\ol H$, making such applications
  impossible. Of course, rather than intertwining the definitions of
  well-formed types of our calculus and the operation of first-order
  substitution on them we could instead prohibit such applications in
  a side condition to the rule for well-formed $\mu$-types, but we
  favor the precision of the current presentation.} We write $\vdash
F$ for $\emptyset;\emptyset \vdash F$.  Definition~\ref{def:wftypes}
ensures that the expected weakening rules for well-formed types hold,
although weakening does not change the contexts in which types can be
formed. If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash
G$, then our rules allow formation of the types $\Gamma;\emptyset
\vdash \Nat^\emptyset F \,G$ and $\Gamma; \emptyset \vdash
\Nat^{\ol\alpha} \,\onet \,F$, which represent the System F types
$\Gamma \vdash F \to G$ and $\forall$-type $\Gamma; \emptyset \vdash
\forall \ol\alpha . F$, respectively. However, some System F types,
such as $\forall \alpha. (\alpha \to \alpha) \to \alpha$, are not
representable in our calculus.

{\color{red} CHANGE TEXT FOR NEW $\mu$-RULE.}
Definition~\ref{def:wftypes} allows the formation of all of the
(closed {\color{red} NOT!}) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
{\color{red} The representation of $\mathit{List}\,\alpha$ as the ADT
  $\mu \beta. \onet + \alpha \times \beta$ cannot be functorial in
  $\alpha$. By contrast, the representation of $\mathit{List}\,\alpha$
  as the nested type $(\mu \phi. \lambda \beta.\,\onet + \beta \times
  \phi \beta)\,\alpha$
%Each of these types
  can be considered either functorial in $\alpha$ or not, according to
  whether $\alpha$ is in the functorial or non-functorial context in
  which this type is formed.} In particular, if $\emptyset;\alpha
\vdash \mathit{List} \,\alpha$, then $\vdash \Nat^\alpha \onet\,
(\mathit{List}\,\alpha)$ is well-formed; if $\alpha;\emptyset \vdash
\List \,\alpha$, then it is not.  Similarly, if
$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. {\color{red}
  KEEP? Although types can be functorial in variables of arity greater
  than $0$, the body $F$ of a type $(\mu \phi.\lambda \ol\alpha.F)\ol
  G$ cannot.} For example, the type $\mathit{GRose}\, \phi \, \alpha =
\mu\beta . \alpha \times \phi \beta$ {\color{red} can be functorial
  --- NOT!} or not in $\alpha$, but cannot be functorial in
$\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot be the
(co)domain of a $\Nat$-type representing a natural transformation
functorial in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
system. These restrictions ensure that the set and relational
interpretations of types in Section~\ref{sec:type-interp} are
functors. In Section~\ref{sec:conclusion} we discuss how more
expressive types might be accommodated.

@@

{\color{red} Check no change needed for new $\mu$-rule in next
  def. Just note that in application case the args to $\phi$ go in for
  the $\alpha$s in $F$s replacing $\phi$.}
\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
It is not hard to see that $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma;
\ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap
\ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.

@@

  \begin{adjustbox}{varwidth=5.4in, max width=5in, fbox, center}
       \[\begin{array}{ccc}
       \AXC{$\Gamma;\Phi \vdash F$}
       \UIC{$\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F$}
       \hspace*{0.2in}\DisplayProof\hspace*{0.05in}
       &
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : \zerot$}
       \AXC{$\Gamma;\Phi \vdash F$}
       \BIC{$\Gamma;\Phi \,|\, \Delta \vdash \bot_F t  : F$}
       \DisplayProof\hspace*{0.05in}
       &
       \AXC{$\phantom{\Gamma;\Phi}$}
       \UIC{$\Gamma;\Phi \,|\, \Delta \vdash \top : \onet$}
       \DisplayProof\\\\
       \end{array}\]

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \Phi \vdash F,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F+G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta, x : F \vdash l : K \hspace{0.2in} \Gamma;\Phi \,|\, \Delta, y : G \vdash r : K$}
       \TIC{$\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} : K$}
       \hspace*{-0.2in}\DisplayProof
       \end{array}\]

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \emptyset
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \emptyset
         ~|~\emptyset
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha}, {\color{red}\ol{\gamma}} \vdash H$}
       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
         \Nat^{\ol{\beta},{\color{red}\ol{\gamma}}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha}, {\color{red} \ol{\gamma}} \vdash H$}
       \AXC{$\Gamma; \ol{\beta}, {\color{red} \ol{\gamma}} \vdash F$}
       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}, {\color{red} \ol{\gamma}}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
          {\color{red} \ol{\gamma}}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

       \caption{Well-formed terms {\color{red} Remove red $\gamma$s?
           Leave rule for map as
           is???}}\label{fig:terms} \vspace*{-0.00in}
\end{adjustbox}
%       \vspace*{-0.25in}
\end{figure*}

@@

{\color{red} Change the rest of the section to agree with new term
  formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
in the rules for $\map^{\ol{F},\ol{G}}_H$ {\color{red} , $\tin_H$, and
  $\fold^F_H$} (i.e., those variables not affected by the substitution
of $\phi$) deserve comment. They allow us to map {\color{red} or fold}
polymorphic functions over nested types. Suppose, for example, we want
to map the polymorphic function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
the absence of extra variables the instance of $\map$ required to map
each non-functorial monomorphic instantiation of $\mathit{flatten}$
over a list of perfect trees is well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
\beta)))\] required to map the {\em polymorphic} $\mathit{flatten}$
function over a list of perfect trees is not: indeed, the functorial
contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
be empty, but because the polymorphic $\mathit{flatten}$ function is
natural in some variable, say $\delta$, it cannot possibly have a type
of the form $\Nat^\emptyset F\, G$ as would be required for it to be
the function input to $\map$. Untypeability of this instance of $\map$
is unsatisfactory in a polymorphic calculus, where we naturally expect
to be able to manipulate entire polymorphic functions rather than just
their monomorphic instances, but the ``extra'' variables $\ol \gamma$
remedy the situation, ensuring that the instance of $\map$ needed to
map the polymorphic $\mathit{flatten}$ function is typeable as
follows:

@@

\noindent
Similar remarks justify the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\mathsf{fold}$.

>>>>

\newcommand{\tb}[1]{~~ \mbox{#1} ~~}
\newcommand{\listt}[1]{(\mu \phi. \lambda \beta . \onet + \beta \times
  \phi \beta) #1} 
\newcommand{\filtype}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (List \, \alpha) \, (List \, \alpha))} 
\newcommand{\filtypeGRose}{\Nat^\emptyset 
 (\Nat^\emptyset \, \alpha \, \mathit{Bool})\, (\Nat^\emptyset 
  (\mathit{GRose}\,\psi \, \alpha) \, (\mathit{GRose}\,\psi \, (\alpha
  + \onet)))} 
\newcommand{\maplist}{\mathit{map}_{\lambda A. \setsem{\emptyset; \alpha
      \vdash \mathit{List} \, \alpha} \rho[\alpha := A]}} 
\newcommand{\PLeaves}{\mathsf{PLeaves}}
\newcommand{\swap}{\mathsf{swap}}
\newcommand{\reverse}{\mathsf{reverse}}
\newcommand{\bcons}{\mathit{bcons}}
\newcommand{\bnil}{\mathit{bnil}}

@@

\begin{defi}\label{def:wftypes}
The formation rules for the set $\F$ of\, {\em well-formed types} are

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[4ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, although weakening does not change the
contexts in which types can be formed. If $\Gamma;\emptyset \vdash F$
and $\Gamma;\emptyset \vdash G$, then our rules allow formation of the
types $\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$ and $\Gamma;
\emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which represent the
System F types $\Gamma \vdash F \to G$ and $\forall$-type $\Gamma;
\emptyset \vdash \forall \ol\alpha . F$, respectively. However, some
System F types, such as $\forall \alpha. (\alpha \to \alpha) \to
\alpha$, are not representable in our calculus. Note that, in the
rule for $\mu$-types, no variables in $\Phi$ are applied to types
involving variables from $\ol\alpha$ in the body $F$ of $\mu
\phi. \lambda \ol\alpha.F$. This will be critical to proving the
Identity Extension Lemma for our calculus.

Definition~\ref{def:wftypes} allows the formation of all of the types
from Section~\ref{sec:intro}:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
\beta. \onet + \alpha \times \beta$ cannot be functorial in
$\alpha$. By contrast, its representation as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
considered either functorial in $\alpha$ or not, according to whether
$\alpha$ is in the functorial or non-functorial context in which this
type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
in $\ol\alpha$. As a result, both $\phi$ and the variables in
$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
cannot be the (co)domain of a $\Nat$-type binding any variables, and
no type of the form $(\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
system. These restrictions ensure that the set and relational
interpretations of types in Section~\ref{sec:type-interp} are
functors. In Section~\ref{sec:conclusion} we discuss how more
expressive types might be accommodated. {\color{red} Daniel: Check the
  above carefully?}

@@

\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, G \,K\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
{\color{red} Daniel: Agree?} The idea is that the arguments to $\phi$
get substituted for the variables in $\ol\alpha$ in each $F$ replacing
an occurrence of $\phi$. It is not hard to see that $\Gamma;\Phi
\vdash H[\phi :=_{\ol{\alpha}} F]$. In addition, if \,$\Gamma, \phi^k;
\Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha|
= k$, and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi
\vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.

@@

  \begin{adjustbox}{varwidth=5.8in, max width=5.5in, fbox, center}
       \[\begin{array}{ccc}
       \AXC{$\Gamma;\Phi \vdash F$}
       \UIC{$\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F$}
       \hspace*{0.2in}\DisplayProof\hspace*{0.05in}
       &
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : \zerot$}
       \AXC{$\Gamma;\Phi \vdash F$}
       \BIC{$\Gamma;\Phi \,|\, \Delta \vdash \bot_F t  : F$}
       \DisplayProof\hspace*{0.05in}
       &
       \AXC{$\phantom{\Gamma;\Phi}$}
       \UIC{$\Gamma;\Phi \,|\, \Delta \vdash \top : \onet$}
       \DisplayProof\\\\
       \end{array}\]

@@

       \[\begin{array}{c}
       \;\;\AXC{$\Gamma; \Phi \vdash F,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : F+G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta, x : F \vdash l : K \hspace{0.2in} \Gamma;\Phi \,|\, \Delta, y : G \vdash r : K$}
       \TIC{$\Gamma;\Phi~|~\Delta \vdash \cse{t}{x \mapsto l}{y \mapsto r} : K$}
       \hspace*{-0.2in}\DisplayProof
       \end{array}\]

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \emptyset
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \emptyset
         ~|~\emptyset
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha} \vdash H$}
       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
         \Nat^{\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha} \vdash H$}
       \AXC{$\Gamma; \ol\beta \vdash F$}
       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
         (\Nat^{\ol{\beta}}\,(\mu \phi.\lambda
         \ol{\alpha}.H)\ol{\beta}\,F)$} 
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

       \caption{Well-formed terms}\label{fig:terms} \vspace*{-0.00in}
\end{adjustbox}
%       \vspace*{-0.25in}
\end{figure*}

@@

The ``extra'' functorial variables $\ol{\gamma}$ in the rules for
$\map^{\ol{F},\ol{G}}_H$ (i.e., those variables not affected by the
substitution of $\phi$) deserve comment. They allow us to map
polymorphic functions over nested types. Suppose, for example, we want
to map the polymorphic function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
the absence of extra variables the instance of $\map$ required to map
each non-functorial monomorphic instantiation of $\mathit{flatten}$
over a list of perfect trees is well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
\beta)))\] required to map the {\em polymorphic} $\mathit{flatten}$
function over a list of perfect trees is not: indeed, the functorial
contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
be empty, but because the polymorphic $\mathit{flatten}$ function is
natural in some variable, say $\delta$, it cannot possibly have a type
of the form $\Nat^\emptyset F\, G$ as would be required for it to be
the function input to $\map$. Untypeability of this instance of $\map$
is unsatisfactory in a polymorphic calculus, where we naturally expect
to be able to manipulate entire polymorphic functions rather than just
their monomorphic instances, but the ``extra'' variables $\ol \gamma$
remedy the situation, ensuring that the instance of $\map$ needed to
map the polymorphic $\mathit{flatten}$ function is typeable as
follows:

@@
